TRIK Studio: Technical Introduction

### 2017-02-06 2017-trikStudioEnglish/trikStudioEnglish.tex
Added a section about parser

Остаток статьи будет построен следующим образом. В главах~\ref{chapter:controlFlowLanguage} и~\ref{chapter:dataFlowLanguage} будут кратко описаны визуальные языки TRIK Studio и их интерпретаторы. Глава~\ref{chapter:commonArchitecture} дает общее представление об архитектуре среды. Дальнейшие главы будут посвящены конкретным подсистемам TRIK Studio. Общие сведения о реализации механизмов коммуникации с роботами даны в главе~\ref{chapter:communications}. Описания интерпретаторов визуальных языков в системе даны в главе~\ref{chapter:interpreters}. Наиболее интересные детали реализации генераторов в текстовые языки из диаграмм потока управления представлены в главе~\ref{chapter:generators}. В главе~\ref{chapter:2dModel} рассказывается об особенностях реализации подсистемы двумерного имитационного моделирования, далее, в главе~\ref{chapter:constraintsChecker} описан язык описания программ автоматической проверки заданий, исполняемых на этой подсистеме. Наконец, заключительная глава подводит итоги работы.

@@

>>>>

Остаток статьи будет построен следующим образом. В главах~\ref{chapter:controlFlowLanguage} и~\ref{chapter:dataFlowLanguage} будут кратко описаны визуальные языки TRIK Studio и их интерпретаторы. Глава~\ref{chapter:commonArchitecture} дает общее представление об архитектуре среды. Дальнейшие главы будут посвящены конкретным подсистемам TRIK Studio. Общие сведения о реализации механизмов коммуникации с роботами даны в главе~\ref{chapter:communications}. Описания интерпретаторов визуальных языков в системе даны в главе~\ref{chapter:interpreters}. Наиболее интересные детали реализации генераторов в текстовые языки из диаграмм потока управления представлены в главе~\ref{chapter:generators}. В разделе~\ref{chapter:parser} описывается подсистема синтаксического анализа текстовой части языка. В главе~\ref{chapter:2dModel} рассказывается об особенностях реализации подсистемы двумерного имитационного моделирования, далее, в главе~\ref{chapter:constraintsChecker} описан язык описания программ автоматической проверки заданий, исполняемых на этой подсистеме. Наконец, заключительная глава подводит итоги работы.

@@

\section{Textual language parser and interpreter}
\label{chapter:parser}
TRIK Studio uses visual languages for robots programming, but arithmetic expressions, intrinsic function calls and so on are better represented with textual strings (in fact, NXT-G tries to use visual blocks even for arithmetic expressions, representing them as syntax trees, it is very inconvenient). As a textual part of a language for both visual languages TRIK Studio uses a subset of Lua 5.3\footnote{https://www.lua.org/}, customized for our needs. Decision to implement our own parser and interpreter of a textual language was made under these considerations:
\begin{itemize}
	\item we needed a small textual language with lightweight syntax and without explicit typing since it shall be used by non-programmers;
	\item we needed explicit abstract syntax tree and the ability to run type inference on it since we were going to generate code on C or EV3 bytecode, which is strongly typed;
	\item we needed to be able to customize language syntax if the need will arise;
	\item all existing implementations for such languages had interpreter without access to a syntax tree or a parser that was not reusable from C++ code;
	\item since we needed only a subset of a language (arithmetic expressions, without statements, custom function and type definitions), writing custom parser was not a difficult task.
\end{itemize}

Parser and interpreter were implemented as a separate library in QReal core, so the resulting textual language is available not only in TRIK Studio, but in all other domain-specific solutions based on QReal. Parser implementation consists of two parts --- general-purpose parser combinator library and Lua parser library implemented using parser combinators. Many projects use ANTLR\footnote{http://www.antlr.org/}, boost.spirit\footnote{http://boost-spirit.com/home/}, yacc\footnote{http://dinosaur.compilertools.net/yacc/} as tools for parser development, but we once again created our custom solution, mainly to avoid additional external dependencies and complication of a build process --- QReal and TRIK Studio are developed by a large community, and not everyone is happy to install additional tools and learn how to use them, especially if their work has nothing to do with syntax analysis and they are students who do not took formal languages course yet.

QReal parser combinator library supports recursive-descent parsers which are able to parse a subset of LL(1) grammars. FOLLOW($\alpha$) set is not calculated, so it limits the expressiveness of resulting grammars, but we were still able to use Lua 5.3 grammar almost as it is specified, with only minor modifications related to factorization and left recursion elimination, which shall be done anyway for LL parsers. For arithmetic expressions Precedence Climbing\footnote{http://www.engr.mun.ca/\~theo/Misc/exp\_parsing.htm} algorithm was used, and it also required some minor alterations in Lua grammar. Custom modifications were made mainly on lexer level to allow, for example, to use '!=' for inequality in addition to '\verb|~=|' used in Lua. Here is a quick example\footnote{full specification of our parser with many irrelevant technical details see at https://github.com/qreal/qreal/tree/master/qrtext/src/lua} of production written in C++ with our library ("`statement is ';' or a list of expressions, optionally followed by '=' and other list of expressions"'):
\begin{verbatim}
	// stat ::= ‘;’ | explist [‘=’ explist]
	stat = (-LuaTokenTypes::semicolon 
	    | (explist & ~(-LuaTokenTypes::equals & explist)))
\end{verbatim}

\verb|LuaTokenTypes::semicolon| is a token corresponding to ';', operator '-' creates a simple parser that can parse only semicolons and excludes semicolon from AST, "`explist"` is a reference parser object, like "`stat"', defined elsewhere, '\&' combines two parsers into a parser which accepts concatenation of their corresponding strings, '|' combines two parsers into a parser that accepts alternative, '\verb|~|' creates optional parser from its argument, which shall also be a parser. There are also operators for adding semantic actions to productions and assigning parser a name for debug purposes. When all parsers are combined in such a way, it is enough to call 'parse()' method of a resulting parser object, giving it a token stream. Parser combinator library was used for another language in QReal~\cite{tikhonova2015generation}, so it is general enough to support not only Lua.

Parser returns abstract syntax tree on which type inference algorithm is executed, providing types for all variables and expressions. Type inference uses Hindley-Milner-style~\cite{damas1982principal} algorithm, simplified for performance reasons and extended to support overloading and coercion. Type inference is also generalised and Lua type inferer only defines inference rules for Lua-specific AST nodes, core type inference functionality is available for all QReal textual languages.

After type inference is complete, expression is ready to be evaluated by an interpreter. Interpreter allows to register intrinsic functions, also it allows to add custom variables with their values, to support using current sensor values in calculations --- robot communication subsystem receives telemetry data from a robot and injects sensor readings into interpreter, which uses them to calculate expressions.

Last notable feature of parsing/interpreting subsystem is extensive use of caching to avoid parsing or reevaluating expressions as much as possible. Program shall be interpreted in real-time, so reparsing and reevaluating expressions every several milliseconds, as required by many control algorithms, would be severe performance problem. But program can be changed by user during interpretation, values for some variables may be changed by external code, such as sensor readings changed by communication subsystem --- so the interpreter keeps track of changes and uses previously calculated values if possible.
